---
title: 'Anaylsis of ATAC-seq in embryos depleted for Dorsal'
author: "Kaelan (1028-02098-001-001)"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    depth: 3
    theme: sandstone
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{1028-02098-001-001}
- \fancyfoot[LE,RO]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options:
  chunk_output_type: console
---

# 1. Introduction

The goal of this analysis is to compare ATAC-seq time course data generated in wild-type, OregonR embryos to the same experiments performed in embryos depleted for Dorsal (gd7). These flies are generated  from a gd7/winscy, P{hs-hid}5 stock that was heat-shocked at the larval stage (see methods). We will use these data to investigate the influence of DV patterning TFs, namely Dorsal, on chromatin accessibility, and will look at this in a pattern-specific way. 

# 2. Computational Setup

```{r, warning=F, message=F}
#Standard packages
library(rtracklayer) ; library(GenomicRanges); library(magrittr) ; library(Biostrings)
library(ggplot2) ; library(reshape2); library(plyranges); library(Rsamtools); library(ggpubr)
library(dplyr); library(data.table); library(patchwork); library(readr); library(testit); library(cowplot)

#KNITR Options
setwd("/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/")
options(knitr.figure_dir="figures/15_dorsal_depletion_atac/", java.parameters = "- Xmx6g")

#Lab sources
source("scripts/r/granges_common.r")
source("scripts/r/metapeak_common.r")
source("scripts/r/knitr_common.r")
source("scripts/r/caching.r")
source("scripts/r/metapeak_functions.r")

#Specific sources
library(ggseqlogo)
library(BSgenome.Dmelanogaster.UCSC.dm6)
library(rhdf5)
source("scripts/r/motif_functions.r")

#Pre-existing variables
modisco_dir <- 'bpnet/modisco/fold1/'
tasks <- c('Zld','Dl','Twi','Bcd','Cad','GAF')
threads <- 5
bsgenome<-BSgenome.Dmelanogaster.UCSC.dm6

```

# 3. Import relevant data for this analysis

```{r import}

# motifs 

motifs.df <- readr::read_tsv('/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/tsv/mapped_motifs/all_instances_curated_1based.tsv.gz') %>% 
  as.data.frame()

# ATAC replicate coverage list: for DESeq2, we will use 3 replicates of all 4 time points, both in wt and gd7 embryos. 

atac.bw.gd7.list <- list(wt_1to15_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_1to15_atac_4_cutsites.bw',
                         wt_1to15_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_1to15_atac_6_cutsites.bw',
                         wt_1to15_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_1to15_atac_7_cutsites.bw',
                         gd7_1to15_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_1to15_atac_3_cutsites.bw',
                         gd7_1to15_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_1to15_atac_5_cutsites.bw',
                         gd7_1to15_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_1to15_atac_6_cutsites.bw',
                         wt_15to2_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_15to2_atac_4_cutsites.bw',
                         wt_15to2_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_15to2_atac_6_cutsites.bw',
                         wt_15to2_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_15to2_atac_7_cutsites.bw',
                         gd7_15to2_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_15to2_atac_3_cutsites.bw',
                         gd7_15to2_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_15to2_atac_5_cutsites.bw',
                         gd7_15to2_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_15to2_atac_6_cutsites.bw',
                         wt_2to25_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_2to25_atac_4_cutsites.bw',
                         wt_2to25_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_2to25_atac_6_cutsites.bw',
                         wt_2to25_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_2to25_atac_7_cutsites.bw',
                         gd7_2to25_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_2to25_atac_3_cutsites.bw',
                         gd7_2to25_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_2to25_atac_5_cutsites.bw',
                         gd7_2to25_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_2to25_atac_6_cutsites.bw',
                         wt_25to3_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_25to3_atac_4_cutsites.bw',
                         wt_25to3_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_25to3_atac_6_cutsites.bw',
                         wt_25to3_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/orer_25to3_atac_7_cutsites.bw',
                         gd7_25to3_rep1 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_25to3_atac_3_cutsites.bw',
                         gd7_25to3_rep2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_25to3_atac_5_cutsites.bw',
                         gd7_25to3_rep3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/individual/gd7_25to3_atac_6_cutsites.bw')

```

# 4. Determine differential accessibility using DESeq2

## 4.1. Format and calculate counts using cutsite information

```{r format}

# Define peaks--these are non-overlapping and evenly-sized

peaks.gr <- rtracklayer::import('/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/narrowpeak/atac_orer_all_peaks.narrowpeak') %>%
  GenomicRanges::reduce(ignore.strand = T) %>%
  GenomicRanges::resize(2114, 'center') %>%
  plyranges::mutate(peak_id = 1:length(.))

# Calculate the counts across all peaks

counts.gd7.df<-mclapply(atac.bw.gd7.list, function(x){
  regionSums(peaks.gr, x)
}, mc.cores = 8) %>% as.data.frame()
colnames(counts.gd7.df)<-names(atac.bw.gd7.list)

# Set conditions for the DESeq2 model

peak_condition_gd7.df<-data.frame(condition = c(rep('wt_1to15', 3), rep('gd7_1to15', 3), rep('wt_15to2', 3), rep('gd7_15to2', 3), rep('wt_2to25', 3), rep('gd7_2to25', 3), rep('wt_25to3', 3), rep('gd7_25to3', 3)))
peak_condition_gd7.df$condition<- factor(peak_condition_gd7.df$condition, levels = c("wt_1to15", "gd7_1to15", "wt_15to2","gd7_15to2", "wt_2to25","gd7_2to25", "wt_25to3","gd7_25to3"))
rownames(peak_condition_gd7.df)<-names(atac.bw.gd7.list)

```

## 4.2. Set up differential model

```{r setup}

library(DESeq2)

dds_gd7 <- DESeqDataSetFromMatrix(countData = counts.gd7.df,
                              colData = peak_condition_gd7.df,
                              design = ~ condition)
model_gd7 <- DESeq(dds_gd7)

```

## 4.3. Gather time point results 

```{r time}

t1_res.gd7.df <- results(model_gd7, contrast = c("condition","gd7_1to15","wt_1to15"), alpha = 0.05)
plotMA(t1_res.gd7.df, ylim=c(-3,3), ylab = 'log fold change log2(gd7_1to15/wt_1to15)') 
t1.ma <- ggplot(as.data.frame(t1_res.gd7.df), aes(x = baseMean, y = log2FoldChange))+
  geom_hline(yintercept = 0, color = 'black', linetype = 'dotted')+
  geom_point(aes(color = padj <0.05), size = .2) +
  #geom_point(color = 'gray', size = .2) +
  scale_y_continuous(limits = c(-3, 3), name = 'log2(gd7_1to15/wt_1to15)')+
  scale_x_continuous(limits = c(0,10000), breaks = c(0,2000,4000,6000,8000,10000))+
  xlab("Mean normalized ATAC-seq signal") +
  ggtitle("MA plot at 1-1.5 hr") +
  theme_cowplot()

t2_res.gd7.df <- results(model_gd7, contrast = c("condition","gd7_15to2","wt_15to2"), alpha = 0.05)
plotMA(t2_res.gd7.df, ylim=c(-3,3), ylab = 'log fold change log2(gd7_15to2/wt_15to2)')
t2.ma <- ggplot(as.data.frame(t2_res.gd7.df), aes(x = baseMean, y = log2FoldChange))+
  geom_hline(yintercept = 0, color = 'black', linetype = 'dotted')+
  geom_point(aes(color = padj <0.05), size = .2) +
  #geom_point(color = 'gray', size = .2) +
  scale_y_continuous(limits = c(-3, 3), name = 'log2(gd7_15to2/wt_15to2)')+
  scale_x_continuous(limits = c(0,10000), breaks = c(0,2000,4000,6000,8000,10000))+
  xlab("Mean normalized ATAC-seq signal") +
  ggtitle("MA plot at 1.5-2 hr") +
  theme_cowplot()

t3_res.gd7.df <- results(model_gd7, contrast = c("condition","gd7_2to25","wt_2to25"), alpha = 0.05)
plotMA(t3_res.gd7.df, ylim=c(-3,3), ylab = 'log fold change log2(gd7_2to25/wt_2to25)')
t3.ma <- ggplot(as.data.frame(t3_res.gd7.df), aes(x = baseMean, y = log2FoldChange))+
  geom_hline(yintercept = 0, color = 'black', linetype = 'dotted')+
  geom_point(aes(color = padj <0.05), size = .2) +
  #geom_point(color = 'gray', size = .2) +
  scale_y_continuous(limits = c(-3, 3), name = 'log2(gd7_2to25/wt_2to25)')+
  scale_x_continuous(limits = c(0,10000), breaks = c(0,2000,4000,6000,8000,10000))+
  xlab("Mean normalized ATAC-seq signal") +
  ggtitle("MA plot at 2-2.5 hr") +
  theme_cowplot()

t4_res.gd7.df <- results(model_gd7, contrast = c("condition","gd7_25to3","wt_25to3"), alpha = 0.05)
plotMA(t4_res.gd7.df, ylim=c(-3,3), ylab = 'log fold change log2(gd7_25to3/wt_25to3)')
t4.ma <- ggplot(as.data.frame(t4_res.gd7.df), aes(x = baseMean, y = log2FoldChange))+
  geom_hline(yintercept = 0, color = 'black', linetype = 'dotted')+
  geom_point(aes(color = padj <0.05), size = .2) +
  #geom_point(color = 'gray', size = .2) +
  scale_y_continuous(limits = c(-3, 3), name = 'log2(gd7_25to3/wt_25to3)')+
  scale_x_continuous(limits = c(0,10000), breaks = c(0,2000,4000,6000,8000,10000))+
  xlab("Mean normalized ATAC-seq signal") +
  ggtitle("MA plot at 2.5-3 hr") +
  theme_cowplot()

maplots <- t1.ma + t2.ma + t3.ma + t4.ma
ggsave("gd7_effect_ma_plots.pdf", plot = maplots, path = "figures/15_dorsal_depletion_atac/", width = 25, height = 20, units = "cm")
ggsave("gd7_effect_ma_plots.png", plot = maplots, path = "figures/15_dorsal_depletion_atac/", width = 25, height = 20, units = "cm")

```

## 4.4. Plot results at enhancers

```{r enhancers}

# Import enhancers  list that contains some DV and AP enhancers that we manually parsed down from larger enhancer lists (Cusanovich, Koenecke, Zeitlinger) for ease of viewing. 

dv_enhancers_subset.gr <- read.csv(file = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/bed/enhancers/dv_enhancers_subset.csv', header = TRUE) %>% makeGRangesFromDataFrame(keep.extra.columns = TRUE)

# Join ATAC peaks with enhancers

peaks.gr$enhancer_name <- NA
enhancer.ov <- findOverlaps(peaks.gr, dv_enhancers_subset.gr, ignore.strand = T)
peaks.gr$enhancer_name[enhancer.ov@from]<-dv_enhancers_subset.gr$name[enhancer.ov@to]

# Add patterning information 

peaks.gr$pattern <- NA
peaks.gr$pattern[enhancer.ov@from]<-dv_enhancers_subset.gr$pattern[enhancer.ov@to]

#Join with DE results (I used the 4th timepoint because this is the peak influence of Dorsal binding, according to Li & Eisen, 2018)

peaks_w_de.df <- cbind(peaks.gr %>% as.data.frame, t4_res.gd7.df) %>%
  dplyr::mutate(has_enhancer = ifelse(is.na(enhancer_name), 'no','yes'))

#Plot with DE results

names_to_show <- c("sog_shadow", "rho", "vn", "brk_primary", "vnd_3", "twi_Ozdemir", "mir-1_Biemar", "Mef2_Nguyen", "sna-S_Perry", "tld_Kirov", "zen_dist_Doyle", "dpp_Huang", "shn", "doc2", "Asph_Ozdemir")

gd7_effect_at_dv_enhancers <- ggplot(peaks_w_de.df, aes(x = baseMean, y = log2FoldChange))+
  geom_hline(yintercept = 0, color = 'black', linetype = 'dotted')+
  geom_point(aes(color = padj <0.05), size = .2) +
  #geom_point(color = 'gray', size = .2) +
  geom_text(data = peaks_w_de.df[peaks_w_de.df$enhancer_name %in% names_to_show,],aes(label = enhancer_name, color = pattern), size = 4)+
  scale_y_continuous(limits = c(-1.5, 1.5), name = 'log2(gd7_25to3/wt_25to3)')+
  scale_x_continuous(limits = c(0,10000), breaks = c(0,2000,4000,6000,8000,10000))+
  xlab("Average of normalized ATAC counts") +
  ggtitle("Gd7 effect at 2.5-3 hr at validated DV patterning enhancers") +
  theme_cowplot()
gd7_effect_at_dv_enhancers

ggsave("gd7_effect_at_dv_enhancers_25to3_w_signif.pdf", plot = gd7_effect_at_dv_enhancers, path = "figures/15_dorsal_depletion_atac/", width = 25, height = 20, units = "cm")
ggsave("gd7_effect_at_dv_enhancers_25to3_w_signif.png", plot = gd7_effect_at_dv_enhancers, path = "figures/15_dorsal_depletion_atac/", width = 25, height = 20, units = "cm")

# Print padj values

peaks_w_de.df[!is.na(peaks_w_de.df$enhancer_name),][peaks_w_de.df[!is.na(peaks_w_de.df$enhancer_name),]$enhancer_name %in% names_to_show, c(7,8,13,14)]

```

## 4.5. Map DESeq2 results back to peaks 

```{r map}

# overlap with peaks

peaks_w_de_t1.df <- cbind(peaks.gr %>% as.data.frame, t1_res.gd7.df) %>%
  dplyr::mutate(has_enhancer = ifelse(is.na(enhancer_name), 'no','yes'))
peaks_w_de_t2.df <- cbind(peaks.gr %>% as.data.frame, t2_res.gd7.df) %>%
  dplyr::mutate(has_enhancer = ifelse(is.na(enhancer_name), 'no','yes'))
peaks_w_de_t3.df <- cbind(peaks.gr %>% as.data.frame, t3_res.gd7.df) %>%
  dplyr::mutate(has_enhancer = ifelse(is.na(enhancer_name), 'no','yes'))
peaks_w_de_t4.df <- cbind(peaks.gr %>% as.data.frame, t4_res.gd7.df) %>%
  dplyr::mutate(has_enhancer = ifelse(is.na(enhancer_name), 'no','yes'))

# make GRanges for export

peaks_w_de_t1.gr <- peaks_w_de_t1.df %>%
  dplyr::mutate(timepoint = '1to15') %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)    

peaks_w_de_t2.gr <- peaks_w_de_t2.df %>%
  dplyr::mutate(timepoint = '15to2') %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

peaks_w_de_t3.gr <- peaks_w_de_t3.df %>%
  dplyr::mutate(timepoint = '2to25') %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

peaks_w_de_t4.gr <- peaks_w_de_t4.df %>%
  dplyr::mutate(timepoint = '25to3') %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

```

## 4.6. Save DESeq2 results

```{r save, eval=FALSE}

deseq_at_peaks.list <- list('gd7_1to15/wt_1to15' = peaks_w_de_t1.gr,
                            'gd7_15to2/wt_15to2' = peaks_w_de_t2.gr,
                            'gd7_2to25/wt_2to25' = peaks_w_de_t3.gr,
                            'gd7_25to3/wt_25to3' = peaks_w_de_t4.gr)
saveRDS(deseq_at_peaks.list, '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/deseq2/gd7/gd7_deseq_timepoints_at_orer_1to3_atac_peaks.RData')

```

## 4.7. Export DESeq2 results as .txt file for GEO

```{r, eval=FALSE}

peaks_w_de_t1.df <- as.data.frame(peaks_w_de_t1.gr) 
peaks_w_de_t1.df$deseq <- "gd7_1to15/wt_1to15"
peaks_w_de_t2.df <- as.data.frame(peaks_w_de_t2.gr)
peaks_w_de_t2.df$deseq <- "gd7_15to2/wt_15to2"
peaks_w_de_t3.df <- as.data.frame(peaks_w_de_t3.gr)
peaks_w_de_t3.df$deseq <- "gd7_2to25/wt_2to25"
peaks_w_de_t4.df <- as.data.frame(peaks_w_de_t4.gr)
peaks_w_de_t4.df$deseq <- "gd7_25to3/wt_25to3"

all_dl_deseq.df <- rbind(peaks_w_de_t1.df,peaks_w_de_t2.df,peaks_w_de_t3.df,peaks_w_de_t4.df)

write.table(all_dl_deseq.df,'/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/deseq2/gd7/gd7_deseq_timepoints_at_orer_1to3_atac_peaks.txt', row.names=FALSE)

```

## 4.8. Map log2FC values onto motifs

```{r motifs}

# Summarize from all time points into one total de onject

peaks_w_de_t1.df$t1_log2FC <- peaks_w_de_t1.df$log2FoldChange
peaks_w_de_t2.df$t2_log2FC <- peaks_w_de_t2.df$log2FoldChange
peaks_w_de_t3.df$t3_log2FC <- peaks_w_de_t3.df$log2FoldChange
peaks_w_de_t4.df$t4_log2FC <- peaks_w_de_t4.df$log2FoldChange

peaks_w_all_time_de.df <- peaks_w_de_t4.df[,1:8]
peaks_w_all_time_de.df$t1_log2FC <- peaks_w_de_t1.df$log2FoldChange
peaks_w_all_time_de.df$t2_log2FC <- peaks_w_de_t2.df$log2FoldChange
peaks_w_all_time_de.df$t3_log2FC <- peaks_w_de_t3.df$log2FoldChange
peaks_w_all_time_de.df$t4_log2FC <- peaks_w_de_t4.df$log2FoldChange

# Map peaks with DESeq information back to motifs 

motifs_vs_peaks.ov<-findOverlaps(motifs.df %>% makeGRangesFromDataFrame(), peaks.gr, ignore.strand = T)
motifs.df$peak_id<-NA
motifs.df$peak_id[motifs_vs_peaks.ov@from]<-peaks.gr$peak_id[motifs_vs_peaks.ov@to]

motifs_across_peaks.df<-motifs.df %>%
  dplyr::filter(!is.na(peak_id)) %>%
  dplyr::left_join(., peaks_w_all_time_de.df %>% dplyr::select(-c(seqnames, start, end, width, strand)), by = 'peak_id') %>%
  as.data.frame()

```

# 5. Compare wt and gd7 ATAC-seq time course data, along with the zld- ATAC, at the patterned enhancers

## 5.1. Import relevant data

```{r relevant}

# Import enhancers from "Genome-wide identification of Drosophila dorso-ventral enhancers by differential histone acetylation analysis" by Koenecke et al., 2016

dv_enhancers.gr <- readRDS('/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/bed/enhancers/koenecke2016_dv_enhancers_denovo_dm6.rds')

# Create coverage bigwig list

atac.total.list <- list(orer_1to1.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/orer_1to15_atac_combined_normalized.bw',
                        orer_1.5to2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/orer_15to2_atac_combined_normalized.bw',
                        orer_2to2.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/orer_2to25_atac_combined_normalized.bw',
                        orer_2.5to3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/orer_25to3_atac_combined_normalized.bw',
                        gd7_1to1.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/gd7_1to15_atac_combined_normalized.bw',
                        gd7_1.5to2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/gd7_15to2_atac_combined_normalized.bw',
                        gd7_2to2.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/gd7_2to25_atac_combined_normalized.bw',
                        gd7_2.5to3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/gd7_25to3_atac_combined_normalized.bw',
                        zldrnai_1to1.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/zldrnai_1to15_atac_combined_normalized.bw',
                        zldrnai_1.5to2 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/zldrnai_15to2_atac_combined_normalized.bw',
                        zldrnai_2to2.5 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/zldrnai_2to25_atac_combined_normalized.bw',
                        zldrnai_2.5to3 = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/data/bw/dm6/atac/combined/zldrnai_25to3_atac_combined_normalized.bw')

```

## 5.2. Calculate ATAC-seq at mesodermal and dorsal ectodermal enhancers

```{r pattern}

atac_across_denovo.df<-mclapply(names(atac.total.list), function(x){
  regionSums(dv_enhancers.gr %>% resize(1000, 'center'), atac.total.list[[x]])
}, mc.cores = 4) %>% as.data.frame()
names(atac_across_denovo.df)<-names(atac.total.list)

dv_enhancers_with_all_atac.df <- cbind(dv_enhancers.gr %>% as.data.frame, atac_across_denovo.df) %>%
  dplyr::select(differential_k27ac, orer_1to1.5, orer_1.5to2, orer_2to2.5, orer_2.5to3, gd7_1to1.5, gd7_1.5to2, gd7_2to2.5,
                gd7_2.5to3, zldrnai_1to1.5, zldrnai_1.5to2, zldrnai_2to2.5, zldrnai_2.5to3) %>%
  as.data.table %>%
  melt.data.table(id.vars = c('differential_k27ac'), variable.name = 'timepoint', value.name = 'ATAC_signal')

```

## 5.3. Plot boxplots for mesodermal vs. dorsal ectodermal enhancers 

```{r boxes}

box_order <- c("orer_1to1.5", "zldrnai_1to1.5", "gd7_1to1.5", "orer_1.5to2", "zldrnai_1.5to2", "gd7_1.5to2", "orer_2to2.5", "zldrnai_2to2.5", "gd7_2to2.5", "orer_2.5to3", "zldrnai_2.5to3", "gd7_2.5to3")
my_comparisons <- list(c("orer_1to1.5", "gd7_1to1.5"), c("orer_1to1.5", "zldrnai_1to1.5"), 
                       c("orer_1.5to2", "gd7_1.5to2"), c("orer_1.5to2", "zldrnai_1.5to2"),
                       c("orer_2to2.5", "gd7_2to2.5"), c("orer_2to2.5", "zldrnai_2to2.5"),
                       c("orer_2.5to3", "gd7_2.5to3"), c("orer_2.5to3", "zldrnai_2.5to3"))
dv_enhancers_with_all_atac.df$timepoint <- factor(dv_enhancers_with_all_atac.df$timepoint, levels = box_order)

# Plot each pattern separately

meso.df <- dv_enhancers_with_all_atac.df[dv_enhancers_with_all_atac.df$differential_k27ac == "Higher in Toll10b",]
dee.df <- dv_enhancers_with_all_atac.df[dv_enhancers_with_all_atac.df$differential_k27ac == "Higher in gd7",]


denovo_meso.plot <- ggplot(meso.df, aes(x=timepoint, y=log2(ATAC_signal))) +
  geom_boxplot(aes(fill = timepoint)) +
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", paired = TRUE, label.y = 13)+
  ylim(8,14) +
  scale_fill_viridis_d()+
  ggtitle("Timecourse ATAC-seq at at mesodermal enhancers") +
  theme_cowplot() +
  theme(legend.position="none")

denovo_dee.plot <- ggplot(dee.df, aes(x=timepoint, y=log2(ATAC_signal))) +
  geom_boxplot(aes(fill = timepoint)) +
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", paired = TRUE, label.y = 13)+
  ylim(8,14) +
  scale_fill_viridis_d()+
  ggtitle("Timecourse ATAC-seq at dorsal ectodermal enhancers") +
  theme_cowplot() +
  theme(legend.position="none")

denovo_boxes <- denovo_dee.plot + denovo_meso.plot + 
  plot_layout(widths = c(2, 2))
denovo_boxes

ggsave("atac_regionsums_at_enhancers_1000window.pdf", plot = denovo_boxes, path = "figures/15_dorsal_depletion_atac/", width = 20, height = 10)
ggsave("atac_regionsums_at_enhancers_1000window.png", plot = denovo_boxes, path = "figures/15_dorsal_depletion_atac/", width = 20, height = 10)

```

## 5.4. Plot ATAC-seq at more limited list with DEE vs. NEE vs. ME information included

```{r neuro}

# Bring in enhancers

dv_w_nee.gr <- read.csv(file = '/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/bed/enhancers/dv_enhancers_w_nee.csv', header = TRUE) %>% makeGRangesFromDataFrame(keep.extra.columns = TRUE)

# Calculate

atac_across_nee.df<-mclapply(names(atac.total.list), function(x){
  regionSums(dv_w_nee.gr %>% resize(1000, 'center'), atac.total.list[[x]])
}, mc.cores = 4) %>% as.data.frame()
names(atac_across_nee.df)<-names(atac.total.list)

dv_enhancers_nee_with_all_atac.df <- cbind(dv_w_nee.gr %>% as.data.frame, atac_across_nee.df) %>%
  dplyr::select(pattern, orer_1to1.5, orer_1.5to2, orer_2to2.5, orer_2.5to3, gd7_1to1.5, gd7_1.5to2, gd7_2to2.5,
                gd7_2.5to3, zldrnai_1to1.5, zldrnai_1.5to2, zldrnai_2to2.5, zldrnai_2.5to3) %>%
  as.data.table %>%
  melt.data.table(id.vars = c('pattern'), variable.name = 'timepoint', value.name = 'ATAC_signal')

# Plot

dv_enhancers_nee_with_all_atac.df$timepoint <- factor(dv_enhancers_nee_with_all_atac.df$timepoint, levels = box_order)

ventral.df <- dv_enhancers_nee_with_all_atac.df[dv_enhancers_nee_with_all_atac.df$pattern == "ventral",]
dorsal.df <- dv_enhancers_nee_with_all_atac.df[dv_enhancers_nee_with_all_atac.df$pattern == "dorsal",]
neuroectoderm.df <- dv_enhancers_nee_with_all_atac.df[dv_enhancers_nee_with_all_atac.df$pattern == "neuroectoderm",]

ventral.plot <- ggplot(ventral.df, aes(x=timepoint, y=log2(ATAC_signal))) +
  geom_boxplot(aes(fill = timepoint)) +
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", paired = TRUE, label.y = 13)+
  ylim(8,14) +
  scale_fill_viridis_d()+
  ggtitle("Timecourse ATAC-seq at mesodermal enhancers") +
  theme_cowplot() +
  theme(legend.position="none")

dorsal.plot <- ggplot(dorsal.df, aes(x=timepoint, y=log2(ATAC_signal))) +
  geom_boxplot(aes(fill = timepoint)) +
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", paired = TRUE, label.y = 13)+
  ylim(8,14) +
  scale_fill_viridis_d()+
  ggtitle("Timecourse ATAC-seq at dorsel ectoderm enhancers") +
  theme_cowplot() +
  theme(legend.position="none")

neuroectoderm.plot <- ggplot(neuroectoderm.df, aes(x=timepoint, y=log2(ATAC_signal))) +
  geom_boxplot(aes(fill = timepoint)) +
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", paired = TRUE, label.y = 13)+
  ylim(8,14) +
  scale_fill_viridis_d()+
  ggtitle("Timecourse ATAC-seq at neuroectoderm enhancers") +
  theme_cowplot() +
  theme(legend.position="none")

patterning_boxes <- dorsal.plot + neuroectoderm.plot + ventral.plot + 
  plot_layout(widths = c(2, 2, 2))
patterning_boxes

ggsave("atac_regionsums_at_patterned_enhancers_1000window.pdf", plot = patterning_boxes, path = "figures/15_dorsal_depletion_atac/", width = 20, height = 10)
ggsave("atac_regionsums_at_patterned_enhancers_1000window.png", plot = patterning_boxes, path = "figures/15_dorsal_depletion_atac/", width = 20, height = 10)

```

# 6. Conclusions

Here, we have characterized the differences in Ore-R and gd7 ATAC-seq and find that the changes depend on the pattern of expression. While mesodermal enhancers, which are Dorsal-activated, decrease in accessibility upon loss of Dorsal, dorsal ectodermal enhancers, which can be Dorsal-repressed, do not lose accessibility but instead can gain it. This suggests that Dorsal's influence on chromatin accessibility is commensurate with its role as an activator or repressor of transcription. 

# 7. Session Information

For the purposes of reproducibility, the R/Bioconductor session information is printed below:

```{r sessioninfo}
sessionInfo()
```
