---
title: 'Plot in silico injections'
author: "Kaelan (1028-02098-001-001)"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    depth: 3
    theme: sandstone
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{1028-02098-001-001}
- \fancyfoot[LE,RO]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options:
  chunk_output_type: console
---

# 1. Introduction

The goal of this analysis is to plot the in silico injections previously generated in 8_acc_insilico_perturbs.ipynb to examine how injecting TF binding motifs into nonpeak regions affects ChromBPNet's chromatin accessibility prediction. Note that this analysis generates the figure 4a. 

# 2. Computational Setup

```{r, warning=F, message=F}
#Standard packages
library(rtracklayer) ; library(GenomicRanges); library(magrittr) ; library(Biostrings)
library(ggplot2) ; library(reshape2); library(plyranges); library(Rsamtools); library(ggpubr)
library(dplyr); library(data.table); library(patchwork); library(readr); library(testit); library(cowplot)

#KNITR Options
setwd("/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/")
options(knitr.figure_dir="figures/8a_plot_insilico_preds/", java.parameters = "- Xmx6g")

#Lab sources
source("scripts/r/granges_common.r")
source("scripts/r/metapeak_common.r")
source("scripts/r/knitr_common.r")
source("scripts/r/caching.r")
source("scripts/r/metapeak_functions.r")

#Specific sources
library(ggseqlogo)
library(BSgenome.Dmelanogaster.UCSC.dm6)
library(rhdf5)
source("scripts/r/motif_functions.r")

#Pre-existing variables
modisco_dir <- 'bpnet/modisco/fold1/'
tasks <- c('Zld','Dl','Twi','Bcd','Cad','GAF')
threads <- 5
bsgenome<-BSgenome.Dmelanogaster.UCSC.dm6

```

# 3. Import and format relevant data

```{r import}

insilico.df<-readr::read_tsv('/l/Zeitlinger/ZeitlingerLab/Manuscripts/Zelda_and_Nucleosomes/Analysis/analysis/tsv/perturbs/accessibility/insilico/insilico_inj_profiles_nonpeak_seq.tsv.gz')

# Re-order columns by motif name and by timepoint

insilico.df$motifA <- factor(insilico.df$motifA, levels = c("Zld", "Zld-lowaff", "Zld-lowestaff", "Gaf", "Cad", "Dl", "Twi", "Bcd"))

insilico.df$timepoint <- factor(insilico.df$timepoint, levels = c("1to15", "15to2", "2to25", "25to3"))

```

# 4. Plot

```{r plot}

insilico.plot <- ggplot(insilico.df, aes(x = position, y = signal, color = type, fill = type))+
  geom_line(size = 1)+
  scale_color_manual(values = c('red3', 'slategray3'), name = '', labels = c("Motif", "No motif"))+
  scale_y_continuous(name = 'Predicted ATAC cuts', limits=c(0, 1.5))+
  scale_x_continuous(name = "Distance from injection (bp)")+
  facet_grid(timepoint ~ motifA) +
  theme_cowplot()
insilico.plot

ggsave("insilico_injections.pdf", plot = insilico.plot, path = "figures/8a_plot_insilico_preds/", width = 60, height = 25, units = "cm")
ggsave("insilico_injections.png", plot = insilico.plot, path = "figures/8a_plot_insilico_preds/", width = 60, height = 25, units = "cm")

```

# 5. Conclusions

Here, we can see that the injection of the Zelda motif and the GAf motif generated chromatin accessibility at all time points, as expected for pioneer TFs. For the following TFs, with the exception of Bcd, we see minor predicted chromatin accessibility in a time point-specific manner (Cad 2-2.5 hr; D; 2.5-3 hr), which is consistent with a more context-specific role of these TFs in accessibility. 

# 6. Session Information

For the purposes of reproducibility, the R/Bioconductor session information is printed below:

```{r sessioninfo}
sessionInfo()
```
